### 2020-06-01 report/sections/introduction/contributions.tex
Report: complete chapter 3 first draft

In \cref{chap:codegen}, we introduce the system architecture for our
code generator, \fancyname{SessionTS}. 
The implementation integrates with the \fancyname{Scribble}
toolchain to generate handler style APIs that the developer can use to write 
their web applications to guarantee protocol conformance by construction.

>>>>

In \cref{chap:codegen}, we introduce the system architecture for our
code generator, \fancyname{SessionTS}. 
The implementation integrates with the Scribble
toolchain to generate handler style APIs that the developer can use to write 
their web applications to guarantee protocol conformance by construction.

### 2020-06-01 report/sections/server/codegen/implementation.tex
Report: progress on chapter 3
Report: complete chapter 3 first draft

\fancyname{SessionTS} is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
We provide a simple command line interface (\cref{lst:cmdline})
for the developer to supply the required information 
to use \fancyname{SessionTS}.

@@

optional arguments:
  -h, --help     show this help message and exit
  -s SERVER, --server SERVER
                 Server role (only applicable for browser targets)
  -o OUTPUT, --output OUTPUT
                 Output directory for generation
\end{lstlisting}
\captionof{figure}{Entry Point for \fancyname{SessionTS}}
\label{lst:cmdline}
\end{figure}

We use Docker \cite{docker} to encapsulate our code generator
and its dependencies -- we found the canonical Python virtual environment
solution to be insufficient, as the Scribble toolchain is
a standalone Java executable with non-trivial setup procedures.
The Dockerfile builds a Docker image with \fancyname{Scribble} and
the Python dependencies all pre-configured, and the provided
\texttt{build.sh} script instantiates the image as a container 
for development.
The \texttt{start.sh} script enters the Docker container
development environment and mounts the local project directory
onto the container as a volume to synchronise changes between the
two environments.

>>>>

\fancyname{SessionTS} is written in Python. 
It offers flexible syntax, a rich standard library, and 
a healthy ecosystem of DOT graph parsers and template engines,
making it a suitable choice for implementing our code generator.
Its rich standard library also simplifies many tasks: we use
the \texttt{argparse} package to generate an informative
command line interface (CLI) for developers to 
supply the correct information (\cref{lst:cmdline})
to use \fancyname{SessionTS},
and \texttt{subprocess} package provides an abstraction to
invoke external tools, such as Scribble and the TypeScript Compiler.

@@

optional arguments:
  -h, --help     show this help message and exit
  -s SERVER, --server SERVER
                 Server role (only applicable for browser targets)
  -o OUTPUT, --output OUTPUT
                 Output directory for generation
\end{lstlisting}
\captionof{lstlisting}{Entry Point for \fancyname{SessionTS}}
\label{lst:cmdline}
\end{figure}

We use Docker \cite{docker} to encapsulate our code generator
and its dependencies -- we found the canonical Python virtual environment
solution to be insufficient, as the Scribble toolchain is
a standalone Java executable with non-trivial setup procedures.
The Dockerfile builds a Docker image with Scribble and
the Python dependencies all pre-configured, and the provided
\texttt{build.sh} script instantiates the image as a container 
for development.
The \texttt{start.sh} script enters the Docker container
development environment and mounts the local project directory
onto the container as a volume to synchronise changes between the
two environments.

### 2020-06-01 report/sections/server/codegen/testing.tex
Report: complete chapter 3 first draft

\section{Testing}
\begin{itemize}
\item unit tests for code generator executable
\item systems tests for type checking the generated output with the typescript compiler
\item sandbox environment set up on the docker container to set up different tsconfig.json files for both react/node targets
\end{itemize}

>>>>

\section{Testing}

The usual unit testing practices from software engineering
apply to our code generator -- the built-in \texttt{unittest} 
package is more than sufficient for our use case. 
Instead, we detail our methodology for \textit{system testing} -- 
executing \fancyname{SessionTS} end-to-end and testing
the generated code.

We implement a test suite on top of \texttt{unittest} APIs
for verifying that \fancyname{SessionTS}
generates \textit{valid TypeScript code}. This is especially useful as
our templates contain both TypeScript syntax and Jinja markup, 
so we cannot easily make sure that each template generates valid code,
let alone checking that the collection of templates generate a valid
TypeScript project altogether.

\begin{figure}[!ht]
\begin{lstlisting}[language=python,tabsize=4]
def test_code_generation(self):
	flags = [scr, protocol, role, target]
	if svr is not None:
		flags.append('-s')
		flags.append(svr)

	rc = mpst_ts.main(flags) (*@\label{line:testmain}@*)
	self.assertEqual(rc, 0)  (*@\label{line:testmainrc}@*)

	completion = subprocess.run(self.npm_test_cmd, shell=True) (*@\label{line:testtsc}@*)
	self.assertEqual(completion.returncode, 0) (*@\label{line:testtscrc}@*)

	shutil.rmtree(self.output_dir)
\end{lstlisting}
\captionof{lstlisting}{Main logic in \fancyname{SessionTS} system testing
test case}
\label{lst:systemtest}
\end{figure}

We provide a collection of Scribble protocols under \texttt{protocols/}
to generate test cases, one per protocol participant.
For each test case, the test suite (\cref{lst:systemtest}) will:

\begin{enumerate}
\item Invoke \fancyname{SessionTS} to generate the TypeScript project 
(\cref{line:testmain}), 
expecting a zero exit code (\cref{line:testmainrc});
\item Run the TypeScript Compiler on the generated directory 
(\cref{line:testtsc}),
passing the \texttt{noEmit} flag to purely perform type-checking,
and expecting a zero exit code (\cref{line:testtscrc}).
\end{enumerate}

As the generated TypeScript code makes assumptions about the
environment in which it is used (for example, having the \texttt{ws}
WebSocket package installed on server-side targets), we require a
sandbox environment to type-check the generated code. 
The sandbox contains the minimal boilerplate required
for testing -- this involves having the WebSocket package installed
for server-side targets, the React.js framework (\cref{section:react})
instantiated for browser-side targets,
and corresponding \texttt{tsconfig.json} files for both targets to
be picked up by the TypeScript Compiler.

For convenience, we extend the
Dockerfile and \texttt{build.sh} script 
to set up the sandbox environments. 
We also make use of the optional
\texttt{--output} flag exposed by the \fancyname{SessionTS} CLI
to redirect the generated code to the correct sandbox environment
to simply the testing process.

### 2020-06-01 report/sections/server/codegen/workflow.tex
Report: complete chapter 3 first draft

We motivate our development workflow from previous work 
\cite{PureScript2019} by extending the \fancyname{Scribble} toolchain
and generating APIs that integrate the developer's 
application logic
into the execution of the communication automata.

@@

\item The developer supplies the communication protocol written in
\fancyname{Scribble} (\cref{subsection:scribble}), 
stating the role (hereafter \textit{endpoint})
to generate APIs for,
and the code generation \textit{target} 
(i.e. whether the role runs on the server or the web browser).

\item \fancyname{SessionTS} delegates to the 
\fancyname{Scribble} toolchain for verifying the well-formedness of
the protocol and expects to receive a DOT graph representation of
the endpoint FSM (\cref{subsection:efsm}). 
\fancyname{SessionTS} parses the endpoint's 
interactions from the DOT graph and generates TypeScript APIs
for the developer (\cref{subsection:apigen}) 
tailored to the specified target.

@@

\subsection{Protocol Specification with \fancyname{Scribble}}
\label{subsection:scribble}

We use the \fancyname{Scribble} protocol description language, 
as presented in
\cite{Scribble}, for formalising the communication structure. This is
inspired by existing work on implementing session type theory 
in mainstream programming languages
\cite{Hybrid2016, PureScript2019, Python2017}. 
We use the variant of the \fancyname{Scribble} language 
previously introduced in \cref{subsection:bgscribble}.

\subparagraph{Type declaration statements}
Specific to our TypeScript API generation toolchain,
the developer is \textit{not} required to explicitly
add type declaration statements for built-in types.
\cref{lst:adder} is a \textit{syntactically correct}
\fancyname{Scribble} protocol as far as 
\fancyname{SessionTS} is concerned. 
Internally, \fancyname{SessionTS} inspects the protocol file
and parses existing type declarations using regular expressions
(or \textit{regex}) -- this is necessary to extract any
custom data types that will appear in the communication (for example,
\cref{lst:game}), and allows \fancyname{SessionTS} to inject
``boilerplate'' type declarations for built-in TypeScript types before
calling \fancyname{Scribble}.

@@

\subsection{From \fancyname{Scribble} to EFSM}
\label{subsection:efsm}

Given the protocol and endpoint, we use \fancyname{Scribble}
to validate the well-formedness of the protocol and extract
information from the protocol relevant for the endpoint.
The latter is expressed as a finite state machine
where each state restricts the possible transitions, 
and transitions between states are represented by
communication actions, i.e. the sending or receiving of a message.

\fancyname{Scribble} expresses the EFSM using the 
DOT graph description language \cite{dot}, with each
communication action encoded as the label of the corresponding
state transition. 
\fancyname{SessionTS} uses the pydot library \cite{pydot}
to parse the graph into an internal representation of the EFSM.
We define an \texttt{EfsmBuilder} class with APIs designed for
constructing the EFSM representation by iterating over the 
state transitions from the DOT representation.

@@

Formally, API generation is a function of the constructed
\texttt{Endpoint} instance
\textit{and} the target specified in the command line. We use a 
different code generation strategy for implementations running on
Node.js versus the browser -- we discuss in greater depth in 
\cref{chap:node} and \cref{chap:react} respectively. 
In this subsection, we explain how we perform API generation
at a higher level of abstraction.

@@

For \fancyname{SessionTS}, we opt to use a template engine for code
generation, namely Jinja \cite{jinja}.

\dots %what is a template, 
\dots %template focuses on how the generated code looks (which makes it easy to debug etc), and uses lightweight markup syntax to inject dynamic content (from our endpoint API)

The main advantage of using a template engine for 
our code generation task is that it decouples the ``presentation''
from the ``content'': in the \fancyname{SessionTS}
% so it is easy to prototype and extend without touching the source code

\begin{itemize}
\item api generation is a function of the EFSM representation
\item traditional methods -- visitor pattern on the EFSM and using stringbuilder to build the file
\item templating library is more suitable to decouple the ``presentation'' (how the code should look) from the ``content'' (the EFSM which is used to build the code)
\item strategy pattern to support the two required build targets (in node and react)
\end{itemize}

>>>>

We motivate our development workflow from previous work 
\cite{PureScript2019} by extending the Scribble toolchain
and generating APIs that integrate the developer's 
application logic
into the execution of the communication automata.

@@

\item The developer supplies the communication protocol written in
Scribble (\cref{subsection:scribble}), 
stating the role (hereafter \textit{endpoint})
to generate APIs for,
and the code generation \textit{target} 
(i.e. whether the role runs on the server or the web browser).

\item \fancyname{SessionTS} delegates to the 
Scribble toolchain for verifying the well-formedness of
the protocol and expects to receive a DOT graph representation of
the endpoint FSM (\cref{subsection:efsm}). 
\fancyname{SessionTS} parses the endpoint's 
interactions from the DOT graph and generates TypeScript APIs
for the developer (\cref{subsection:apigen}) 
tailored to the specified target.

@@

\subsection{Protocol Specification with Scribble}
\label{subsection:scribble}

We use the Scribble protocol description language, 
as presented in
\cite{Scribble}, for formalising the communication structure. This is
inspired by existing work on implementing session type theory 
in mainstream programming languages
\cite{Hybrid2016, PureScript2019, Python2017}. 
We use the variant of the Scribble language 
previously introduced in \cref{subsection:bgscribble}.

\subparagraph{Type declaration statements}
Specific to our TypeScript API generation toolchain,
the developer is \textit{not} required to explicitly
add type declaration statements for built-in types.
\cref{lst:adder} is a \textit{syntactically correct}
Scribble protocol as far as 
\fancyname{SessionTS} is concerned. 
Internally, \fancyname{SessionTS} inspects the protocol file
and parses existing type declarations using regular expressions
(or \textit{regex}) -- this is necessary to extract any
custom data types that will appear in the communication (for example,
\cref{lst:game}), and allows \fancyname{SessionTS} to inject
``boilerplate'' type declarations for built-in TypeScript types before
calling Scribble.

@@

\subsection{From Scribble to EFSM}
\label{subsection:efsm}

Given the protocol and endpoint, we use Scribble
to validate the well-formedness of the protocol and extract
information from the protocol relevant for the endpoint.
The latter is expressed as a finite state machine
where each state restricts the possible transitions, 
and transitions between states are represented by
communication actions, i.e. the sending or receiving of a message.

Scribble expresses the EFSM using the 
DOT graph description language \cite{dot}, with each
communication action encoded as the label of the corresponding
state transition. 
\fancyname{SessionTS} uses the pydot library \cite{pydot}
to parse the graph into an internal representation of the EFSM.
We define an \texttt{EfsmBuilder} class with APIs designed for
constructing the EFSM representation by iterating over the 
state transitions from the DOT representation.

@@

Formally, API generation is a function of the constructed
\texttt{Endpoint} instance
\textit{and} the target specified in the command line. We use a 
different code generation strategy for implementations running on
the servre (\cref{chap:node}) versus the browser (\cref{chap:react}).
In this subsection, we explain how we perform API generation
at a higher level of abstraction.

@@

For \fancyname{SessionTS}, we leverage the Jinja \cite{jinja} 
template engine library for code generation. 
We first construct templates for the TypeScript files we wish to generate,
specifying placeholders for dynamic content (to be extracted
from the \texttt{Endpoint} object); 
we then provide Jinja with the template path and the 
\texttt{Endpoint} object, and the template engine renders the
TypeScript code by filling in the dynamic placeholders. 
We show an example in \cref{lst:jinja}.

\begin{figure}[!ht]
\begin{lstlisting}[language=javascript, title=efsm.ts.j2]
export namespace Message {
{% for state in endpoint.efsm.nonterminal_states %}
  {% for action in state.actions -%}
  export type S{{ state ~ action.label }} = {
    label: Labels.S{{ state }}.{{ action.label }},
    payload: [{{ action.payloads|join(', ') }}],
  };
  {% endfor %}
  export type S{{ state }} = {% for action in state.actions -%}
  | S{{ state ~ action.label }}{% endfor %};
{% endfor %}
\end{lstlisting}
\captionof{lstlisting}{Example Jinja Template for
\fancyname{SessionTS} API Generation}
\label{lst:jinja}
\end{figure}

Jinja provides lightweight syntax for injecting content and
markup for simple control structure: \texttt{\{\{ state \}\}} 
denotes a placeholder for Jinja to render the \texttt{state} variable,
and the \texttt{\{\% \%\}} syntax  is used for conditionals and
control structures 
(such as for loops, to dynamically render the enclosing ``sub-template''
by iterating over a collection).
The main advantage that Jinja brings is that 
it decouples the ``presentation''
from the ``content'' and makes it quick to prototype and extend
the generated code, \textit{usually} without modifications to the 
code generator.

\begin{figure}[!ht]
\begin{lstlisting}[language=python]
class CodeGenerationStrategy(ABC):

    target_to_strategy = {} (*@\label{line:mapping}@*)

    def __init__(self):
        super().__init__()

    @classmethod
    def __init_subclass__(cls, *, target):
        CodeGenerationStrategy.target_to_strategy[target] = cls (*@\label{line:register}@*)
        return super().__init_subclass__()

    @abstractmethod
    def generate(self, endpoint: Endpoint):
        pass
\end{lstlisting}
\captionof{lstlisting}{Implementing \texttt{CodeGenerationStrategy}}
\label{lst:codegenerationstrategy}
\end{figure}

As we generate a different set of TypeScript artefacts depending
on the specified target, we structure the different code generators
using the Strategy design pattern. Each target extends the
abstract base class \texttt{CodeGenerationStrategy} 
and implements its own \texttt{generate()}
method to return a list of \texttt{(path, content)} tuples. 
We define a \texttt{CodeGenerator} class that is parameterised 
by \texttt{target}: when instantiated, it will select and perform 
the specialised \texttt{generate()} method based on \texttt{target}, 
before formatting and committing the generated code 
to the file system.
We implement a subclass hook in \texttt{CodeGenerationStrategy}
(\cref{lst:codegenerationstrategy}),
such that each derived class must provide the target name to
``register'' with the base class (\cref{line:register}),
and the base class keeps an internal
mapping of the concrete strategies (\cref{line:mapping}); 
\texttt{CodeGenerator} accesses
this mapping to select the appropriate strategy.
